---
title: "Figure 6. Source data and GraphPad Prism project inventory"
author: "Reproducibility documentation"
date: "2026-08-27"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

required_packages <- c("knitr", "readxl")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages)) {
  stop("Install the following R packages before knitting: ", paste(missing_packages, collapse = ", "))
}
```

## Overview

This document inventories the statistical source data and native GraphPad Prism project files supplied for Figure 6. The statistical graphs and analyses were created in GraphPad Prism; therefore, this document does not recalculate statistical tests or regenerate the graphs in R.

The source package covers panels A, D, E, G, H, I, J, N, O, and P.

## Source file manifest

```{r source-file-manifest}
manifest <- read.csv("fig6_source_file_manifest.csv", stringsAsFactors = FALSE, check.names = FALSE)
knitr::kable(manifest)
```

## Panel and endpoint mapping

```{r panel-endpoint-mapping}
dictionary <- read.csv("fig6_data_dictionary.csv", stringsAsFactors = FALSE, check.names = FALSE)
knitr::kable(dictionary)
```

## Integrity checks

```{r integrity-checks}
expected_sheets <- c("6A", "6D", "6E", "6G", "6H", "6I", "6J", "6N", "6O", "6P")
observed_sheets <- readxl::excel_sheets("fig6_data.xlsx")
stopifnot(identical(observed_sheets, expected_sheets))

expected_prism_files <- file.path("prism", paste0("fig", expected_sheets, ".pzfx"))
stopifnot(all(file.exists(expected_prism_files)))
stopifnot(file.exists("fig6.pdf"))

prism_headers <- vapply(expected_prism_files, function(path) {
  paste(readLines(path, n = 2L, warn = FALSE), collapse = " ")
}, character(1))
stopifnot(all(grepl("GraphPadPrismFile", prism_headers, fixed = TRUE)))

data.frame(
  item = c("Excel worksheets", "Prism projects", "Figure PDF"),
  observed = c(length(observed_sheets), length(expected_prism_files), 1L),
  expected = c(length(expected_sheets), length(expected_prism_files), 1L),
  status = "verified"
)
```

## Excel source tables

```{r workbook-summary}
sheet_summary <- do.call(rbind, lapply(observed_sheets, function(sheet_name) {
  x <- readxl::read_excel("fig6_data.xlsx", sheet = sheet_name, col_names = FALSE)
  data.frame(
    panel = sheet_name,
    rows = nrow(x),
    columns = ncol(x),
    nonempty_cells = sum(!is.na(as.matrix(x)))
  )
}))
knitr::kable(sheet_summary)
```

The tables below reproduce the cell values stored in the Excel source workbook without changing the original workbook.

```{r workbook-tables, results='asis'}
for (sheet_name in observed_sheets) {
  x <- readxl::read_excel("fig6_data.xlsx", sheet = sheet_name, col_names = FALSE)
  cat("\n\n### Panel ", substring(sheet_name, 2), "\n\n", sep = "")
  print(knitr::kable(x, col.names = NULL))
}
```

## Labeling note

The source files retain their original labels. Some Prism projects use `NC` instead of `CN` and `FF` instead of `OSS`. The final figure terminology is documented in `fig6_data_dictionary.csv`; no source values or labels were modified.

## Reproducibility information

```{r reproducibility-information}
sessionInfo()
```
